
\documentclass[]{article}
\usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage[thmmarks]{ntheorem}
  \usepackage{xspace,cleveref,xcolor}
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{graphicx,url}
  \usepackage[font=small]{caption}
 \usepackage{subcaption}
 \usepackage[parfill]{parskip}
 \usepackage{newclude}
\usepackage{listings}
\usepackage{float}
\usepackage[all]{xy}
\usepackage{tikz}
\usepackage{hhline}

% new commands
  \newcommand{\QQ}{\mathbb Q}
  \newcommand{\RR}{\mathbb R}
  \newcommand{\CC}{\mathbb C}
  \newcommand{\NN}{\mathbb N}
  \newcommand{\ZZ}{\mathbb Z}
  \newcommand{\DD}{\mathbb D}
  \newcommand{\A}{\mathcal A}
  \newcommand{\B}{\mathcal B}
  \newcommand{\C}{\mathcal C}
  \newcommand{\D}{\mathcal D}
  \newcommand{\R}{\mathcal R}
  \newcommand{\M}{\mathcal M}
  \newcommand{\laplace}{\operatorname\Delta}
  \bibliographystyle{alpha}

  \newcommand{\p}{\ensuremath{\mathcal P}\xspace}
  \newcommand{\np}{\ensuremath{\mathcal{NP}}\xspace}
  \newcommand{\fp}{\ensuremath{\mathcal{FP}}\xspace}
  \newcommand{\abs}[1]{\left|#1\right|}
  \newcommand{\sharpp}{\ensuremath{\# \mathcal{P}}\xspace}
  \newcommand{\code}{\texttt}
  \newcommand{\cc}{\code{C++}\xspace}
  \newcommand{\ccx}{\code{C++11}\xspace}
  \newcommand{\irram}{\code{iRRAM}\xspace}
  \newcommand{\MPFR}{\code{MPFR}\xspace}
  \newcommand{\baana}{\code{BA\_ANA}\xspace}
  \newcommand{\anarect}{\code{ANA\_RECT}\xspace}
  \newcommand{\powerseries}{\code{POWERSERIES}\xspace}
  \newcommand{\poly}{\code{POLY}\xspace}
  \newcommand{\func}{\code{FUNC}\xspace}
  \newcommand{\real}{\code{REAL}\xspace}
  \newcommand{\complex}{\code{COMPLEX}\xspace}
  \newcommand{\temp}{\textcolor{red}}
  \newcommand{\seq}{\mathbf}
  \DeclareMathOperator{\lb}{lb}
  \DeclareMathOperator{\bigo}{O}
  \newcommand{\sprec}{prec\xspace}
  \newcommand{\demph}{\textbf}
  \newcommand{\sdzero}{\texttt{0}}
% ntheorem environments
  \renewenvironment{abstract}{%
    \hfill\begin{minipage}{0.95\textwidth}
      \small \textbf{\abstractname.}}
      {\end{minipage}}
  \theoremseparator{:}
  \theorembodyfont{\itshape}
  \newtheorem{definition}{Definition}[section]
  \theorembodyfont{\upshape}
  \newtheorem{theorem}[definition]{Theorem}
  \newtheorem{corollary}[definition]{Corollary}
  \newtheorem{example}[definition]{Example}
  \newenvironment{proof}{\paragraph{Proof:}}{\hfill$\square$}
  \newenvironment{proofsketch}{\paragraph{Proof (Sketch):}}{\hfill$\square$}
\lstset{language={C++},
                basicstyle=\ttfamily,
                keywordstyle=\color{red}\ttfamily,
                stringstyle=\color{blue}\ttfamily,
                commentstyle=\color{green}\ttfamily,
                morecomment=[l][\color{magenta}]{\#}
}
\title{Case Studies in Exact Real Arithmetic - Implementations and empirical
Evaluations}
\author{Holger Thies}
\date{}
\begin{document}
\maketitle
The most common way to represent real numbers on a computer is the use of floating point
arithmetic.
The floating point representation of a real number is a finite word, consisting
of a mantissa and an exponent of fixed length.
Due to the finiteness of the representation, it is unavoidable that errors
occur.
While in many cases the error stays small, there are also several examples
where computations with floating point numbers yield results that are
completely wrong and far off from the correct value.

In any case, the error is not visible from the representation, thus without
additional analysis it is impossible to know if the error is too big for a
particular application or not.
Floating point arithmetic is therefore not a reliable way to perform real number
computations.

This thesis deals with computations on real numbers that are reliable in the
sense that there are algorithms that can compute the result with guaranteed
user-defined error bounds.
This is known as \textbf{exact real arithmetic}.

Computable Analysis gives a sound theoretical foundations for such
computations.
In Computable Analysis, a real number $x$ is called \textbf{computable} if there is
an algorithm (e.g. formalized via Turing-machines) that outputs on input $n$ a rational number $q$ such that
$\abs{x-q} \leq 2^{-n}$. 
By this definition, it is obvious that uncomputable real numbers exist, since
the set of real numbers is uncountable, while the set of algorithms is
countable.

The so called Type-2 Theory of Effectivity (TTE) \cite{Wei} extends this definition
further, to make it possible to talk about computability and complexity of more
general continuous objects like real functions.

There already exists a vast theory of computability and complexity built on
TTE.
It is an important task to compare the claims made in theory to actual
implementations in exact real arithmetic.

Algorithms for exact real arithmetic can be implemented in modern programming
languages and there are already some frameworks that provide functionality to
simplify such computations.

This thesis uses Norbert M\"{u}ller's \cc package \irram \cite{irram} for implementations.
\irram extends \cc by classes and functions for error-free computations with
real numbers.
The most important class is the class \real. An object of type \real is meant
to behave like a  real number that can be manipulated without rounding errors.
Objects of type \real can for example be multiplied, added or subtracted.
An approximation to a real number represented by such an object can also be
printed with any desired precision.
It is guaranteed that the error of the approximation will be small enough to
ensure said precision.
The \irram package takes care of the steps necessary for this. 
It does so by iteratively increasing the precision of the computations.
That is, in a computation instead of a real number $x$ an interval $I = [d-e,
d+e]$ with $x \in I$ is used.
Interval arithmetic is used to bound the error on the result of the performed
operations.
If an approximation to a real number is needed at some point and the precision
does not suffice, the whole computation is started from the beginning with
higher precision, i.e., the intervals are made smaller. 
This process is repeated until the precision is sufficient.

The thesis first gives an introduction to the theory of computability and
complexity on real numbers and an overview of ways to deal with real numbers on
a computer with a detailed view on the \irram package.

It then proceeds by presenting two case studies in exact real arithmetic, i.e.,
two numerical problems are considered in detail from the viewpoint of real
complexity theory and implemented in \cc using the \irram package. 
The results from theory are compared with time measurements obtained from
empirical evaluation.

The first case-study shows how exact real arithmetic can be used to get simpler
algorithms for classical problems in numerical analysis.
The particular problem considered is to compute the shadowing distance for a chaotic
dynamic systems.
The simplest case of a discrete dynamical system is when the state space is 1-dimensional (say $X \subseteq \RR$) and 
the transition function only depends on the previous value,
formally $\Phi(1,x) = f(x)$ for some $f : \RR \to \RR$. 
In this case it can be written by the recurrence relation $x_{n+1} = f(x_n)$ with initial condition $x_0 \in \RR$.
The set of the $x_n$ is called \textbf{orbit} of the map $f$.

When computing an orbit of a chaotic system numerically, i.e., using floating-point
approximations, the computed orbit is usually far off from the real (exactly
computed) orbit. 
A shadowing orbit is a real orbit that stays close to a numerically computed
orbit.
The case study empirically checks how long such shadowing orbits exist
depending on the starting point and on the approximation precision and how
close the numerical and the shadowing orbit are.
The chaotic system considered is the logistic map.

This problem was originally considered by Hammel, Yorke and Grebogi in 1987
\cite{Hammel1987}. 
Since they could not compute real numbers exactly, they used a form of interval
arithmetic to bound the exact orbit.
However, \irram can be used to compute the exact orbit which thus yields a
simplified version of their algorithm.
Additionally a multi-precision arithmetic version is presented using
\code{MPFR}.

The second case-study deals with analytic functions and is based on the work 
of Kawamura, M\"{u}ller, R\"{o}snick and Ziegler \cite{Kawamura}.
Analytic functions have been thoroughly studied in real complexity theory as a
subset of real functions where many in general computationally hard problems become feasible. 
An analytic function is locally defined by its Taylor series.
It has been shown that a function $f: \RR \to \RR$ is computable if and only if its Taylor
series is computable \cite{pour1989computability}.
The same holds for polynomial time computability \cite{mueller1987uniform}.
Many operations on functions, like addition, multiplication, differentiation or
anti-differentiation, can be reduced to simple manipulations on the Taylor
series. Thus, those operations can easily seen to be polynomial time computable for polynomial
time computable analytic functions.

However, the above theorems are non-uniform in the sense that 
they only say that $f$ being polynomial time computable implies the existence of a polynomial time computable 
coefficient sequence and the existence of such a sequence implies that there exists some algorithm computing $f$ in polynomial time.
In no way, however, do they say, how to compute the sequence from a given representation of the function and vice versa.
In fact it can be shown that when given only the Taylor series, not even evaluation of the corresponding function is computable  \cite{mueller1987uniform}. 
For uniform computability, some additional information has to be provided.
The thesis discusses two possible ways to represent analytic functions that yield uniform polynomial
time computability of many important operations.
The first is for functions $f \in C^\omega(\overline{B_r(x_0)})$, i.e., functions analytic on the closed disc of radius r around $x_0$.
The additional information needed in this case are two integer parameters $k$ and $A$, such that
$\sqrt[k]{2} \leq r$ and $|a_j|r^j \leq A$ for all $j \in \NN$ where $r = (\limsup |a_j|^{\frac{1}{j}})^{-1}$ denotes the radius of convergence of the series.

More general domains can be covered by balls and for each such ball a series and parameters as above can be given. 
This is, however, cumbersome to find for complicated domains. 
For functions analytic on a real line $[a,b]$ it was thus also considered to choose parameters $A$ and $k$ that hold globally for all the series used to cover the domain and only save one of the coefficient series.
The other series can be found by analytic continuation, which has been shown to be polynomial time computable as well  \cite{Muller1993}. 
It should be noted however, that normally analytic continuation has to be iterated several times until the right series is computed and that this process is not polynomial time in the number of iterations.

The above has been implemented using the \irram package.
The implementation is meant as an extension to \irram, that
provides user-friendly classes for computations with analytic functions.
The implementation consists of two main classes, \baana and \anarect that are used for functions analytic on $B_r(x_0)$ 
and $[a,b]$, respectively.

Both classes provide a common set of operators and methods, in particular
\begin{enumerate}
  \item It is possible to add, multiply and subtract two objects using the overloaded operators $+$, $*$, $-$.
  \item It is possible to evaluate the function at $x \in D$ using the overloaded operator $()$
  \item There is a function \code{differentiate (unsigned int n)} to get a new object representing the $d$-th derivative of the analytic function.
  \item There is a function \code{integrate (unsigned int n)} to get a new object representing the $d$-th anti-derivative of the analytic function.
  \item It is possible to get the coefficients of the underlying Taylor series.
\end{enumerate}

Empirical evaluation was performed on the running time of those classes and compared
with the expected running times from the theoretical examination.
The results of this evaluation coincide well with the theoretical running time bounds.

\bibliography{thesis}
\end{document}
