\documentclass[tudarticle,type=msc,colorback,accentcolor=tud9c, twoside]{tudthesis}
\usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage[thmmarks]{ntheorem}
  \usepackage{xspace,cleveref,xcolor}
  \usepackage{amsmath}
  \usepackage{graphicx,url}
  \usepackage[font=small]{caption}
 \usepackage{subcaption}
 \usepackage[parfill]{parskip}
 \usepackage{newclude}
\usepackage{listings}
\usepackage{float}
\usepackage[all]{xy}
\usepackage{tikz}
\usepackage{hhline}

% new commands
  \newcommand{\QQ}{\mathbb Q}
  \newcommand{\RR}{\mathbb R}
  \newcommand{\CC}{\mathbb C}
  \newcommand{\NN}{\mathbb N}
  \newcommand{\ZZ}{\mathbb Z}
  \newcommand{\DD}{\mathbb D}
  \newcommand{\A}{\mathcal A}
  \newcommand{\B}{\mathcal B}
  \newcommand{\C}{\mathcal C}
  \newcommand{\D}{\mathcal D}
  \newcommand{\R}{\mathcal R}
  \newcommand{\M}{\mathcal M}
  \newcommand{\laplace}{\operatorname\Delta}
  \bibliographystyle{alpha}

  \newcommand{\p}{\ensuremath{\mathcal P}\xspace}
  \newcommand{\np}{\ensuremath{\mathcal{NP}}\xspace}
  \newcommand{\fp}{\ensuremath{\mathcal{FP}}\xspace}
  \newcommand{\abs}[1]{\left|#1\right|}
  \newcommand{\sharpp}{\ensuremath{\# \mathcal{P}}\xspace}
  \newcommand{\code}{\texttt}
  \newcommand{\cc}{\code{C++}\xspace}
  \newcommand{\ccx}{\code{C++11}\xspace}
  \newcommand{\irram}{\code{iRRAM}\xspace}
  \newcommand{\MPFR}{\code{MPFR}\xspace}
  \newcommand{\baana}{\code{BA\_ANA}\xspace}
  \newcommand{\anarect}{\code{ANA\_RECT}\xspace}
  \newcommand{\powerseries}{\code{POWERSERIES}\xspace}
  \newcommand{\poly}{\code{POLY}\xspace}
  \newcommand{\func}{\code{FUNC}\xspace}
  \newcommand{\real}{\code{REAL}\xspace}
  \newcommand{\complex}{\code{COMPLEX}\xspace}
  \newcommand{\temp}{\textcolor{red}}
  \newcommand{\seq}{\mathbf}
  \DeclareMathOperator{\lb}{lb}
  \DeclareMathOperator{\bigo}{O}
  \newcommand{\sprec}{prec\xspace}
  \newcommand{\demph}{\textbf}
  \newcommand{\sdzero}{\texttt{0}}
% ntheorem environments
  \renewenvironment{abstract}{%
    \hfill\begin{minipage}{0.95\textwidth}
      \small \textbf{\abstractname.}}
      {\end{minipage}}
  \theoremseparator{:}
  \theorembodyfont{\itshape}
  \newtheorem{definition}{Definition}[section]
  \theorembodyfont{\upshape}
  \newtheorem{theorem}[definition]{Theorem}
  \newtheorem{corollary}[definition]{Corollary}
  \newtheorem{example}[definition]{Example}
  \newenvironment{proof}{\paragraph{Proof:}}{\hfill$\square$}
  \newenvironment{proofsketch}{\paragraph{Proof (Sketch):}}{\hfill$\square$}
\lstset{language={C++},
                basicstyle=\ttfamily,
                keywordstyle=\color{red}\ttfamily,
                stringstyle=\color{blue}\ttfamily,
                commentstyle=\color{green}\ttfamily,
                morecomment=[l][\color{magenta}]{\#}
}
\begin{document}
  \thesistitle{Case Studies in Exact Real Arithmetic - Implementations and empirical Evaluation}{Fallstudien in exakter reeller Arithmetik: Implementation und empirische Evaluation}
  \author{Holger Thies}
  \referee{Prof. Dr. Martin Ziegler}{}
  \department{Department of Mathematics}
  \group{Mathematical Logic Group}
  \dateofexam{\today}{\today}
  \makethesistitle
  \affidavit{H. Thies}
\begin{abstract}
 The thesis gives examples for the possibilities and limitations of exact real
 arithmetic in form of two case studies.
 Theoretical results from real complexity theory are compared with
 time measurements obtained from empirical evaluation of algorithms implemented
 in the \irram
 \cc framework for exact real arithmetic.
 An overview on some results from computable analysis and an introduction to
 the \irram framework is included.
\end{abstract}  
\vfill
  \dedication{
    \textbf{Acknowledgments.} I first want to thank my supervisor, Martin Ziegler, for giving me the
    opportunity to work on this topic and the helpful advice he provided. 
    I would also like to thank him for his kind and constant support during my graduate
    and undergraduate education.

    I also want to thank Christoph Spandl and Peter Hertling from the
    Bundeswehr University Munich for providing the topic for the
    second case study.

    Further, I would like to thank Akitoshi Kawamura, under whose supervision a
    large part of the second case study was conducted.
    I am also very grateful to him and Hiroshi Imai for allowing me to stay at
    the University of Tokyo as an exchange student and for their support during
    this stay.

    Last but not least, I want to thank Florian Steinberg, for working together
    with me on the second case study. 
    Many of the algorithms were implemented by him and he had lots of great
    ideas on how to improve the implementation.
  }
\tableofcontents
  \chapter{Introduction}
  \include*{tex/introduction/introduction}
  \chapter{Theoretical Background}
  \include*{tex/computability/computability}
  \include*{tex/complexity/complexity}
  \chapter{Computing with Reals}
  \include*{tex/float/float}
  \include*{tex/multiprec/multiprec}
  \include*{tex/interval/interval}
  \include*{tex/symbolic/symbolic}
  \include*{tex/exact/exact}
  \include*{tex/irram/irram}
  \chapter{Case Study: Dynamic Systems and the Shadowing Lemma}
  \include*{tex/dynamic_systems/dynamic_systems}
  \chapter{Case Study: A Datatype for Analytic Functions}
  \include*{tex/analytic_functions/analytic}
  % \chapter{Case Study: Root Refinement for Real Polynomials}
  % \section{Introduction}
  % \section{Implementation}
  % \section{Evaluation}
  \chapter{Conclusion}
  \include*{tex/conclusion/conclusion}
  \bibliography{thesis}
\end{document}
